Predicting the spread and persistence of genetically modified dominant sterile male mosquitoes

### 2021-01-14 inst/manuscript/Draft/bmc_article_v9.tex
Rename draft dir, changed prior script

\begin{abstract} % abstract
\parttitle{Background} Reproductive containment provides an opportunity to implement a staged-release strategy for genetic control of malaria vectors, in particular allowing predictions about the spread and persistence of (self-limiting) sterile and male-biased strains to be compared to outcomes before moving to (self-sustaining) gene-drive strains.
\parttitle{Methods} We: (i) describe a diffusion-advection-reaction model of the spread and persistence of a single cohort of male mosquitoes; (ii) elicit informative prior distributions for model parameters, for wild type and genetically modified dominant sterile strains; (iii) estimate posterior distributions for wild type strains using data from published Mark Recapture Release experiments, with inference performed through a delayed rejection adaptive Metropolis algorithm; and, (iv) weight prior distributions in order to make predictions about genetically modified strains using Bayes factors calculated for the wild type strains.
\parttitle{Results} If a single cohort of 5000 genetically modified dominant sterile male mosquitoes are released at the same location as the previous MRR experiments, there is a 90\% probability that the expected number of released mosquitoes will fall below 1 in 10 days, and that by 12 days there will be a 99\% probability that no mosquitoes will be found more than 150 meters from the release location.
\parttitle{Conclusions} Spread and persistence models should form a key component of risk assessments of novel genetic control strategies for malaria vectors. Our predictions, used in an independent risk assessment, suggest genetically modified sterile male mosquitoes will remain within the locality of the release site, and persist for a very limited amount of time. Data gathered following the release of these mosquitoes will enable us to test the accuracy of these predictions and also provide a means to update parameter distributions for genetic strains in a coherent (Bayesian) framework. We anticipate this will provide additional insights about how to conduct probabilistic risk assessments for the stage-released of genetically modified mosquitoes.
\end{abstract}

@@

In this analysis, we assume that village compounds provide a source of CO$_2$ that acts as an attractant for female \emph{Anopheles gambiae sensu strictu} and \emph{Anopheles coluzzi} mosquitoes \cite{DeJong1995, Majeed2017}. We further assume that male mosquitoes swarm in places where the probability of encountering receptive females is highest \citep{Diabate2015}, hence swarms occur in and around compounds (as evident in Figure 1 of \citep[][]{Epopa2017}). Consequently, high concentrations of attractant result in flights that (on average) are closer to the direction of the nearest compound for both males and females. Conversely, low concentrations have less influence on the mosquitoes and result in greater variability in flight direction. We do not incorporate the effect of wind speed or direction into the analysis (we assume the local dispersal dynamics are not greatly affected by the wind at this spatial scale), and also assume that the attraction process is not influenced by the number of mosquitoes.

@@

\subsection{Data}
%\paragraph{MRR data - what parameters is the data talking to}
In this analysis we use the results of five MRR experiments \cite{Epopa2017} wherein approximately 5000 adult male WT mosquitoes, marked with a colored powder, were released two hours before swarming (around 4pm), at three different locations, on five different dates (Table~\ref{tab:relLoc}). Mosquitoes were recaptured in three ways via: (i) swarm sampling in the evening (times were not specified, but we assume the sampling occurred between 7pm and 9pm); (ii) Pyrethroid Spray Catches in the morning inside houses; and, (iii) placement of humidified clay pots within rooms within houses, that were subsequently checked in the morning (between 6am-7am).

@@

The model's two chemotaxis-related parameters are: (i) the relative strength of attraction towards the source of the attractant that a mosquito experiences at a location ($\va$), with units $m^{-1} R^{-1}$ where $m$ is the distances in meters between the location and the source and $R$ is a measure of the ``reward'' at the source, often expressed as the concentration of $CO_2$ for female mosquitoes; and, (ii) the distance beyond which a mosquito is negligibly attracted to a swarm site, called the range ($\sigma$) with units $m$. The attractiveness of a swarm site at distance $\sigma$ from a mosquito is roughly 38\% of its attractiveness at zero distance.

@@

\subsection{Spread and persistence of male DSM mosquitoes}
The predicted persistence of DSM mosquitoes following a release of a single cohort of 5000 males is summarised in Figure \ref{fig:plotTimeMoxMaxLoc}. The large uncertainty captured in Table \ref{tble:sumStatPPpostbma} is clearly reflected in these predictions: the mean expected survival is estimated to be about 9 to 10 days, with a 90\% probability that the expected number of male DSM mosquitoes falls below 1 by Day 10. The 90\% credible interval, however, is large, and there exists a very low probability that it could take as long about two months for this to occur.

@@

The dispersal of the DSM mosquitoes is balanced between the fairly large values of the dispersion parameter and the fairly high mortality rate. Because of this, the DSM are not really expected to disperse far from their release site, with the expected number of mosquitoes per cell delineated by the red and orange contours in Figure \ref{fig:plotSpread}. For instance, while the spread is expected to extend on average to up to $500$m away from the release location by day $2$ (with $1\%$ chance that it extends further), the extent then contracts quickly to a limited area (less than $250$m from release location) by day $5$, to finally about $150$m away from the release location at day $12$.

@@

This staged-learning is borne out by this analysis: the reproductive containment strategy utilised in the DSM mosquitoes provides a good, relatively simple, starting point for this type of analysis. The reaction dynamics in our model are greatly simplified by virtue of the fact that the released mosquitoes are sterile. The dynamic models for the second stage in Target Malaria's pathway, namely self-limiting male bias strains, will be more complicated because they must accommodate the birth processes and genetic dynamics that are ignored here, and also other potentially important inter-specific interactions \citep{Beeton2020}. The models for the third and final stage, that is self-sustaining driving male bias strains, will be more complicated again particularly because the inter-specific interactions are likely to be more important and because the spatio-temporal scope of the analysis will be significantly larger.

>>>>

\begin{abstract} % abstract
\parttitle{Background} Reproductive containment provides an opportunity to implement a staged-release strategy for genetic control of malaria vectors, in particular allowing predictions about the spread and persistence of (self-limiting) sterile and male-biased strains to be compared to outcomes before moving to (self-sustaining) gene-drive strains.
\parttitle{Methods} We: (i) describe a diffusion-advection-reaction model of the spread and persistence of a single cohort of male mosquitoes; (ii) elicit informative prior distributions for model parameters, for wild type (WT) and genetically modified dominant sterile strains(DSM); (iii) estimate posterior distributions for wild type strains using data from published Mark Recapture Release (MRR) experiments, with inference performed through a delayed rejection adaptive Metropolis algorithm; and, (iv) weight prior distributions in order to make predictions about genetically modified strains using Bayes factors calculated for the wild type strains.
\parttitle{Results} If a single cohort of 5000 genetically modified dominant sterile male mosquitoes are released at the same location as the previous MRR experiments with WT, there is a 90\% probability that the expected number of released mosquitoes will fall below 1 in 10 days, and that by 12 days there will be a 99\% probability that no mosquitoes will be found more than 150 meters from the release location.
\parttitle{Conclusions} Spread and persistence models should form a key component of risk assessments of novel genetic control strategies for malaria vectors. Our predictions, used in an independent risk assessment, suggest genetically modified sterile male mosquitoes will remain within the locality of the release site, and persist for a very limited amount of time. Data gathered following the release of these mosquitoes will enable us to test the accuracy of these predictions and also provide a means to update parameter distributions for genetic strains in a coherent (Bayesian) framework. We anticipate this will provide additional insights about how to conduct probabilistic risk assessments for the stage-released of genetically modified mosquitoes.
\end{abstract}

@@

In this analysis, we assume that village compounds (houses) provide a source of CO$_2$ that acts as an attractant for female \emph{Anopheles gambiae sensu strictu} and \emph{Anopheles coluzzi} mosquitoes \cite{DeJong1995, Majeed2017}. We further assume that male mosquitoes swarm in places where the probability of encountering receptive females is highest \citep{Diabate2015}, hence swarms occur in and around compounds (as evident in Figure 1 of \citep[][]{Epopa2017}). Consequently, high concentrations of attractant result in flights that (on average) are closer to the direction of the nearest compound for both males and females. Conversely, low concentrations have less influence on the mosquitoes and result in greater variability in flight direction. We do not incorporate the effect of wind speed or direction into the analysis (we assume the local dispersal dynamics are not greatly affected by the wind at this spatial scale), and also assume that the attraction process is not influenced by the number of mosquitoes.

@@

\subsection{Data}
%\paragraph{MRR data - what parameters is the data talking to}
In this analysis we use the results of five MRR experiments \cite{Epopa2017} wherein approximately 5000 adult male WT mosquitoes, marked with a colored powder, were released two hours before swarming (around 4pm), at three different locations, on five different dates (Table~\ref{tab:relLoc}). Mosquitoes were recaptured in three ways via: (i) swarm sampling in the evening (times were not specified, but we assume the sampling occurred between 7pm and 9pm); (ii) Pyrethroid Spray Catches in the morning inside houses; and, (iii) placement of humidified clay pots within rooms within houses, that were subsequently checked in the morning (between 6am-7am). All releases and re-captures sites were geo-located allowing for a spatial model to be used.

@@

The model's two chemotaxis-related parameters are: (i) the relative strength of attraction towards the source of the attractant that a mosquito experiences at a location ($\va$), with units $m^{-1} R^{-1}$ where $m$ is the distance in meters between the location and the source and $R$ is a measure of the ``reward'' at the source, often expressed as the concentration of $CO_2$ for female mosquitoes; and, (ii) the distance beyond which a mosquito is negligibly attracted to a swarm site, called the range ($\sigma$) with units $m$. The attractiveness of a swarm site at distance $\sigma$ from a mosquito is roughly 38\% of its attractiveness at zero distance.

@@

\subsection{Spread and persistence of male DSM mosquitoes}
The predicted persistence of DSM mosquitoes following a release of a single cohort of 5000 males is summarised in Figure \ref{fig:plotTimeMoxMaxLoc}. The large uncertainty captured in Table \ref{tble:sumStatPPpostbma} is clearly reflected in these predictions: the mean expected survival is estimated to be about 9 to 10 days, with a 90\% probability that the expected number of male DSM mosquitoes falls below 1 by Day 10. The 90\% credible interval, however, is large, and there exists a very low probability ($\sim 0.05$) that it could take as long about two months for this to occur.

@@

The dispersal of the DSM mosquitoes is balanced by the mortality rate. Because of this, the DSM are not really expected to disperse far from their release site, with the expected number of mosquitoes per cell delineated by the red and orange contours in Figure \ref{fig:plotSpread}. For instance, while the spread is expected to extend on average to up to $500$m away from the release location by day $2$ (with $1\%$ chance that it extends further), the extent then contracts quickly to a limited area (less than $250$m from release location) by day $5$, to finally about $150$m away from the release location at day $12$.

@@

This staged-learning is borne out by this analysis: the reproductive containment strategy utilised in the DSM mosquitoes provides a good, relatively simple, starting point for this type of analysis. The reaction dynamics in our model are greatly simplified by virtue of the fact that the released mosquitoes are sterile. The dynamic models for the second stage in Target Malaria's pathway, namely self-limiting male bias strains, will be more complicated because they must accommodate the birth processes and genetic dynamics that are not relevant here, and also other potentially important inter-specific interactions \citep{Beeton2020}. The models for the third and final stage, that is self-sustaining driving male bias strains, will be more complicated again particularly because the inter-specific interactions are likely to be more important and because the spatio-temporal scope of the analysis will be significantly larger.
